---
title: "Overview of R Package hdmax2"
author: "Basile Jumentier, Claire-Cécile Barrot, Olivier Francois"
date: "September 8, 2022"
output:
  prettydoc::html_pretty:
    self_contained: true
    theme: cayman
    highlight: github
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

***

**Summary:**

***

## Introduction 

HDMAX2 is an efficient algorithm for high dimensional mediation analysis using max-squared tests, considering CpGs and aggregated mediator regions. HDMAX2 is structured in three main steps.

- Step 1. **Evaluating associations between exposure, mediators and outcome** using latent factor mixed models (LFMMs) to estimate the effects of exposure $X$ on a matrix $M$ of CpG markers, and the effect of each marker on outcome, $Y$. The models can be described as follows

$$
 M=Xa^T +U_1V_1^T +E_1
$$
where $a$ is the vector of effect sizes of exposure on DNAm levels, $U$ a matrix formed of $K$ latent factors estimated simultaneously with $a$, $V$ are loadings associated with the latent factors, and  $E_1$ is a matrix of residual errors (EWAS 1),

$$
Y=Xc+Mb^T +U_2V_2^T +E_2 
$$
where is $c$ the direct effect of exposure on outcome, $b$ the effect sizes of DNAm levels on outcome, $U_2$ the latent factors from a latent factor regression model, $V_2$ the corresponding loadings, and$E_2$ the a matrix of errors (EWAS 2).

- Step 2. **Identifying potential CpG mediators and aggregated mediator regions**. This step combines the significance values $P_x$ and $P_y$ obtained in EWAS 1 and 2 of step 1 to compute a single  $P$-value for each CpG marker.

$$
P = \max(P_x, P_y)^2
$$
to complete. 

- Step 3. **Quantifying indirect effects with single and multiple mediators**.
to complete.


***
## Starting with hdmax2

To install the latest version of **hdmax2**, use the github repository 
```{r}
devtools::install_github("bcm-uga/hdmax2")
```
The hdmax2 package can be loaded as follows. 
```{r}
library(hdmax2)
```

In this tutorial, we present an example of mediation analysis. The example file includes randomized human DNA methylation values, with simulated treatment (exposure), simulated phenotype (outcome), and simulated covariables.  

```{r}
load("example_hdmax2.data")
```
***

## Evaluating associations between exposure, mediators and outcome

In order to estimate the number of latent factors in the LFMMs, we perform a principal component analysis (PCA) from the methylation matrix using the **prcomp** function as follows. 

```{r}
pc <- prcomp(methylation)
plot(pc$sdev[1:15]^2,
     xlab = 'Principal Component',
     ylab = "Explained variance",
     col = c(rep(1, 3), 2, rep(1, 16)))
```

The screeplot indicates around 4 main components in the data. We will use $K = 4$ latent factors in subsequent analyses. The **mEWAS** function is then applied to estimate the effects of exposure $X$ on a matrix $M$ of CpG markers, and the effect of each marker on outcome $Y$. It uses some observed covariables ($conf$) and $K = 4$ latent factors.

```{r}
##Epigenome Wide Association Study with both exposure and outcome
res.mEWAQ <- mEWAS(X = exposure,
                   Y = phenotype,
                   M = methylation,
                   K = 4,
                   conf = covariables) 
```

The function estimates the effects of $X$ and $Y$ on the matrix $M$. It also estimates the  loadings for the four latent factors.

```{r}
##P-values of X and Y
pvalues <- res.mEWAQ$pValue

##scores matrix for the K latent factors
score_latent <- res.mEWAQ$U

```

The function **mEWAS** also computes the Genomic Inflation Factor (gif), which represents the deviation of the distribution of the observed test statistic compared to the distribution of the expected test statistic.

```{r}
##Genomic inflation factor for X and Y
res.mEWAQ$gif
```

In case of highly stratified population (high gif), The gif is used to recalibrate $P$-values.

```{r}
##Calibrated P-values of X and Y
pvalues.cal <- res.mEWAQ$calibrated.pvalue

##Calibrated scores
score_latent.cal <- res.mEWAQ$calibrated.score2
```
***
## CpG mediators

**Identifying potential CpG mediators**

Combining the significance values $pvalues$ with the function **max2**. This test rejects the null-hypothesis that either the effect of exposure on DNAm or the effect of DNAm on outcome is null.

```{r}
cpg_max2 <- max2(pval1 = pvalues[, 1], pval2 = pvalues[, 2], diagnostic.plot = T)
```

The function **max2** computes a P-value and estimates a Q-value for every CpG. It quantifies the local false discovery rate (FDR) parameter eta0.

```{r}
##local FDR parameter
cpg_max2$eta0
```

In this example, we will consider FDR levels lower than 10%.


```{r}
##Selecting CpG
cpg <- cpg_max2$qval[cpg_max2$qval<= 0.1]
cpg
```

**Quantifying indirect effects of single mediator**

The function **wrap_mediation** estimates the mediation effect sizes for each CpG.
```{r}
effect_cpg <- wrap_mediation(qval = cpg_max2$qval,
                            X = exposure,
                            Y = phenotype,
                            M = methylation,
                            U = res.mEWAQ$U,
                            sims = 100, FDR = 0.1)

##The parameter sims is the number of Monte Carlo draws for nonparametric bootstrap or quasi-Bayesian approximation. 10000 is recommended, but we used 100 for the example.


```

The average causal mediation effect (ACME) is the indirect effect. It also evaluates the average direct effect (ADE), the proportion mediated (PM), the total effect, and the regressions of X on M and M on Y.

```{r}
effect_cpg$ACME
```
***
## Identifying aggregated mediator regions (AMR)

**Identifying AMR**

The function **AMR_search** will identify aggregated methylated regions (AMR) using the max2 P-values. It computes the P-value and the FDR for each AMR.

```{r}
##Detecting AMR
res.amr_search <- AMR_search(chr = annotation$chr,
                   start = annotation$start,
                   end = annotation$end,
                   pval = cpg_max2$pval,
                   cpg = annotation$cpg)
res.amr_search$res
```
The function **AMR_build** builds  a vector for each AMR, containing details for the AMR.

```{r}
res.arm_build <- AMR_build(res.amr_search, methylation = methylation, nb_cpg = 2)
##List of DMR
res.arm_build$res
```
```{r}
##CpG in the DMR
res.arm_build$CpG_for_each_AMR
```

**Quantifying indirect effects**

The function **wrap_mediation_AMR** is similar to the function **wrap_mediation**.
```{r}
effect_amr <- wrap_mediation_AMR(X = exposure,
                            Y = phenotype,
                            AMR = res.arm_build$AMR_mean,
                            U = res.mEWAQ$U,
                            sims = 100)
```

```{r}
##Indirect effect
effect_amr$ACME
```
```{r}
##Direct effect
effect_amr$ADE
```
```{r}
##Peoportion mediated
effect_amr$PM
```
```{r}
##Total effect
effect_amr$TE
```
```{r}
##regressions of X on the DMR
effect_amr$xm
```
```{r}
##regressions of the DMR on Y
effect_amr$my
```






